Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit

### 2024-06-03 optim.qmd
example 2 is there :)

```{r setup, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
```

@@

## Another data example

[TODO: overall summary]construct a relationship between jellyfish success and jellyfish parameters and the optimisation properties defined in @sec-theory. This can inform the choice of jellyfish parameters for a given
optimisation problem.

In addition to the pipe-finding problem, the sine-wave finding problem is also investigated in 6D and 8D spaces. Six indices are considered: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`, which gives the 12 problems printed as rows in @tbl-smoothness-squintability. Combining these problems with different jellyfish optimiser parameters (`n_jellies` and `max_tries`), a total of 51 observations is produced, comprising of 24 observations for the pipe-finding and 27 observations for the sine-wave finding.

For each observation, fifty (50) runs of the jellyfish optimisation is performed. In each run, summary statistics are calculated to find the best index across all the jellyfishes (`I_max`) and the proportion of jellyfishes found (`P_J`). At the observation level, the best index value across all the 50 runs  (`I_max_max`) and the proportion of trials found (`P_J_hat`) is calculated.

Smoothness and squintability measures are computed for each problem, following the procedures outlined in @sec-smoothness and @sec-squintability. To calculate smoothness, 300 random bases are used to estimate Gaussian process parameters (variance, range, smoothness, and nugget). For squintability, 50 random bases are sampled and interpolated to the optimal basis using a step size of 0.005. The index value and projection distance are calculated for all the interpolated bases. A binning procedure is then applied to average the index value at each bin width of 0.005. Finally, the scaled logistic function with four parameters (theta1 - theta4) is estimated by non-linear least square to obtain the squintability measure.

@tbl-smoothness-squintability presents all the parameters calculated for smoothness and squintability, with the column "smooth" for smoothness and "theta3" for squintability. @fig-idx-proj-dist shows the relationship between the index value against the projection distance for each problem following the binning procedure.

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (including variance, range, smoothness, and nugget) and scaled logistic function (theta1 - theta4) for both the pipe-finding and sine-wave finding problems. The "smooth" and "theta3" column represent the smoothness and squintability measures. 
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
library(kableExtra)
tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  knitr::kable(digits = 3) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(3, 7, 11), border_right = TRUE) |> 
  kable_styling(latex_options = "hold_position")
```

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values against projection distance for the 12 pipe/sine-wave finding problem after the binning procedure during the estimation of the squintability measure. The index value is averaged at each bin width of 0.005 and the TIC index is scaled to 0-1 for comparison. When finding the sine-wave structure using the MIC and TIC index, a convex curved is observed, as opposed to the pipe-finding problem or the sine-wave finding problem with the dcor2d\\_2, loess2d, and splines2d index. When finding the sine-wave with the stringy index, the index shows an instantaneous jump to the optimum when close to the best basis."
load(here::here("data/sq_basis_dist_idx.rda"))
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = idx, group = index)) +
  geom_line() +
  facet_nested_wrap(~index + n, nrow = 3, labeller = label_both, scales = "free_y") + 
  xlab("projection distance") + ylab("index value")

@@

```{r eval = FALSE}
#| label: tbl-all-sim
#| tbl-cap: The first five rows of the datasets after combining all the calculation. The smoothness and squintability variable are uniquely characterised by the index function used and the data dimension, and thus do not vary across n\\_jellies and max\\_tries. Variables I\\_max\\_max, P\\_J\\_hat, and time are calculated at each 51 observations.
load(here::here("data/sim_df.rda"))
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                I_max_max, P_J_hat, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(5) |> 
  kable(digits = 3)
```

>>>>

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

## Connecting optimisation success with its properties and jellyfish parameters

The optimisation properties (smoothness, squintability, and speed) proposed in the @sec-theory provide numerical measures to characterise the difficulty of the optimisation for a given projection pursuit problem. This example investigates how these measures, combined with the jellyfish parameters, affect the success of the jellyfish optimisers and how they can guide the decisions made in hyperparameter tuning with the jellyfish optimiser. Simulations are conducted with different jellyfish optimisers to obtain the proportion of successful optimisation across a collection of scenarios, for which smoothness, squintability, and speed are calculated. A generalised linear model is used to link the proportion of success rate with the jellyfish parameters and the optimisation properties.

In addition to the pipe-finding problem, also investigated is the sine-wave finding problem in 6D and 8D spaces with six indices considered: `dcor2d_2`, `loess2d`, `MIC`, `TIC`, `spline`, and `stringy`. Combining with different jellyfish optimiser parameters (`n_jellies` and `max_tries`), a total of `r nrow(sim_df)` cases is produced, comprising of `r nrow(filter(sim_df, index == "holes"))` pipe-finding cases  and `r nrow(filter(sim_df, index != "holes"))` sine-wave finding cases. For each case, the jellyfish optimiser is run fifty (50) times and summary statistics are calculated on the best index value found across all 50 runs and the proportion of runs that find a close (with a difference less than 0.05) best index value (`P_J_hat`).

Smoothness and squintability are computed for each case, following the procedures outlined in @sec-smoothness and @sec-squintability. To calculate smoothness, three hundred random bases are simulated. Index values and projection distance to the optimal basis are calculated for each basis to fit a Gaussian process model. For squintability, fifty random bases are sampled and interpolated to the optimal basis with a step size of 0.005. The interpolated bases are first binned with a bin width of 0.005 to average the index values before feeding into a 4-parameter scaled logistic function, estimated by non-linear least square, to obtain the squintability measure.

@tbl-smoothness-squintability presents the parameters estimated for each case from the Gaussian process (variance, range, smooth, and nugget) for smoothness and the scaled logistic function (theta1 to theta4, and squint) for squintability. The column "smooth" and "squint" are used as the smoothness and squintability measure. The low squint value of MIC and TIC index is due to its convex shape of the index value against the projection distance, as shown in @fig-idx-proj-dist. Comparing to other concave shapes, the maximum first derivative happens at a smaller projection distance ($\theta2$), requiring the optimiser to get closer to the optimal to see a significant change in the index value, hence a small squintability measure. [TODO: the stringy squintability doens't make sense]

@tbl-mod-data presents the processed data for modelling after augmenting the jellyfish parameters (`n_jellies` and `max_tries`) and the average time spent on each run (`time`) for each case. A generalised linear model is fitted with a binomial family and a logit link function to the data to investigate the relationship between the proportion of jellyfish success and jellyfish parameters and the optimisation properties. Pre-processing of the data include 1) scale the jellyfish parameters (`n_jellies` and `max_tries`) by 10 for interpretation, and 2) create a binary variable `long_time` to flag the cases with average time spent on each run greater than 30 seconds. @tbl-mod-output presents the model output with the estimated coefficients. The model suggests a positive relationship between the proportion of jellyfish success and the jellyfish parameters, `n_jellies` and `max_tries`, optimisation property parameter, `smoothness` and `squintability`, and a negative relationship with the average time spent on each run (`long_time`) and the data dimension (`d`). The variable `squintability` and `d` are significant, suggesting their importance over jellyfish parameters to the success of the optimisation.

@@

```{r}
#| label: tbl-smoothness-squintability
#| tbl-cap: Parameters estimated from the Gaussian process (including variance, range, smoothness, and nugget) and scaled logistic function (theta1 to theta4) for the pipe-finding and sine-wave finding problems. The squint column is calculated as $\frac{\theta1 \theta2 \theta3}{4}$, as described in @sec-squintability. The "smooth" and "squint" column represent the smoothness and squintability measures. 
tibble(shape = c(rep("pipe", 4), rep("sine", 8)), smoothness) |>
  left_join(squintability) |> 
  rename(d = n, smooth = smoothness) |> 
  knitr::kable(digits = 3, booktabs = T) |> 
  column_spec(1, border_left = TRUE) |> 
  column_spec(c(3, 7, 12), border_right = TRUE) |> 
  kable_styling(latex_options = "hold_position", font_size = 7)
```

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values against projection distance for the 12 pipe/sine-wave finding problem after the binning procedure during the estimation of the squintability measure. The index value is averaged at each bin width of 0.005 and the `TIC` index is scaled to 0-1 for comparison. When finding the sine-wave structure using the `MIC` and `TIC` index, a convex curved is observed, as opposed to the pipe-finding problem or the sine-wave finding problem with the `dcor2d_2`, `loess2d`, and `splines2d` index. When finding the sine-wave with the stringy index, the index shows an instantaneous jump to the optimum when close to the best basis."
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = idx, group = index)) +
  geom_line() +
  facet_nested_wrap(~index + n, nrow = 3, labeller = label_both, scales = "free_y") + 
  xlab("projection distance") + ylab("index value")

@@

```{r}
#| label: tbl-mod-data
#| tbl-cap: The first few rows of the datasets processed for modelling. The smoothness and squintability variable are uniquely characterised by the index function used and the data dimension, and thus do not vary across `n_jellies` and `max_tries`. The variable `P_J_hat`, and time are calculated at each observation.
sim_df |> 
  dplyr::select(index, d, smoothness, squintability, n_jellies, max_tries, 
                -I_max_max, P_J_hat, time) |> 
  arrange(index, d, n_jellies, max_tries) |> 
  head(7) |> 
  kable(digits = 3)
```

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success (`P_J_hat`) on optimisation properties (`smoothness`, `squintability`, and `d`) and jellyfish parameters (`long_time`, `n_jellies`, and `max_tries`) from the generalised linear model with a binomial family and a logit link function. The variable `long_time` is derived to flag the cases with average time spent on each run greater than 30 seconds and the variable `n_jellies` and `max_tries` are scaled by 10 for interpretation.
sim_df2 <- sim_df |> 
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 30, 1, 0),
         P_J_hat = ifelse(index == "stringy", 0, P_J_hat))

mod1 <- glm(P_J_hat ~ smoothness + squintability + d + long_time + n_jellies + max_tries, data = sim_df2, family = binomial)
broom::tidy(mod1) |> 
  mutate(term = ifelse(term == "(Intercept)", "intercept", term)) |>
  kable(digits = 3) |> 
  column_spec(c(1,5), border_right = TRUE) |> 
  column_spec(1, border_left = TRUE)
```

### 2024-06-03 optim.tex
example 2 is there :)

\begin{verbatim}
Warning: package 'ggplot2' was built under R version 4.3.2
\end{verbatim}

\begin{verbatim}
Warning: package 'tidyr' was built under R version 4.3.2
\end{verbatim}

\begin{verbatim}
Warning: package 'ggh4x' was built under R version 4.3.2
\end{verbatim}

@@

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, leftrule=.75mm, arc=.35mm, left=2mm, opacityback=0, breakable, toptitle=1mm, toprule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, colback=white, titlerule=0mm, bottomrule=.15mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame]

@@

\subsection{Another data example}\label{another-data-example}

{[}TODO: overall summary{]}construct a relationship between jellyfish
success and jellyfish parameters and the optimisation properties defined
in Section~\ref{sec-theory}. This can inform the choice of jellyfish
parameters for a given optimisation problem.

In addition to the pipe-finding problem, the sine-wave finding problem
is also investigated in 6D and 8D spaces. Six indices are considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}, which gives the 12 problems
printed as rows in Table~\ref{tbl-smoothness-squintability}. Combining
these problems with different jellyfish optimiser parameters
(\texttt{n\_jellies} and \texttt{max\_tries}), a total of 51
observations is produced, comprising of 24 observations for the
pipe-finding and 27 observations for the sine-wave finding.

For each observation, fifty (50) runs of the jellyfish optimisation is
performed. In each run, summary statistics are calculated to find the
best index across all the jellyfishes (\texttt{I\_max}) and the
proportion of jellyfishes found (\texttt{P\_J}). At the observation
level, the best index value across all the 50 runs
(\texttt{I\_max\_max}) and the proportion of trials found
(\texttt{P\_J\_hat}) is calculated.

Smoothness and squintability measures are computed for each problem,
following the procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, 300 random
bases are used to estimate Gaussian process parameters (variance, range,
smoothness, and nugget). For squintability, 50 random bases are sampled
and interpolated to the optimal basis using a step size of 0.005. The
index value and projection distance are calculated for all the
interpolated bases. A binning procedure is then applied to average the
index value at each bin width of 0.005. Finally, the scaled logistic
function with four parameters (theta1 - theta4) is estimated by
non-linear least square to obtain the squintability measure.

{[}TODO: discuss what the values of smooth and theta3 mean for the
optimisation{]}

Table~\ref{tbl-smoothness-squintability} presents all the parameters
calculated for smoothness and squintability, with the column ``smooth''
for smoothness and ``theta3'' for squintability.
Figure~\ref{fig-idx-proj-dist} shows the relationship between the index
value against the projection distance for each problem following the
binning procedure.

@@

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (theta1 - theta4) for the pipe-finding and
sine-wave finding problems. The squint column is calculated as theta1 *
theta2 * theta3 / 4, as described in Section~\ref{sec-squintability}.
The ``smooth'' and ``squint'' column represent the smoothness and
squintability measures.}

@@

@@

\caption{\label{fig-idx-proj-dist}Index values against projection
distance for the 12 pipe/sine-wave finding problem after the binning
procedure during the estimation of the squintability measure. The index
value is averaged at each bin width of 0.005 and the TIC index is scaled
to 0-1 for comparison. When finding the sine-wave structure using the
MIC and TIC index, a convex curved is observed, as opposed to the
pipe-finding problem or the sine-wave finding problem with the
dcor2d\_2, loess2d, and splines2d index. When finding the sine-wave with
the stringy index, the index shows an instantaneous jump to the optimum
when close to the best basis.}

@@

\caption{\label{tbl-all-sim}The first few rows of the datasets processed
for modelling. The smoothness and squintability variable are uniquely
characterised by the index function used and the data dimension, and
thus do not vary across n\_jellies and max\_tries. The variable
P\_J\_hat, and time are calculated at each observation.}

@@

\begin{verbatim}
# A tibble: 7 x 5
  term          estimate std.error statistic p.value
  <chr>            <dbl>     <dbl>     <dbl>   <dbl>
1 (Intercept)     -4.67      4.51     -1.04  0.300  
2 smoothness       1.60      1.53      1.05  0.294  
3 squintability    7.06      2.21      3.20  0.00140
4 d               -0.595     0.260    -2.29  0.0221 
5 long_time       -0.851     1.35     -0.632 0.528  
6 n_jellies        0.230     0.131     1.75  0.0795 
7 max_tries        0.107     0.153     0.700 0.484  
\end{verbatim}

>>>>

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, coltitle=black, colframe=quarto-callout-note-color-frame, opacityback=0, toprule=.15mm, opacitybacktitle=0.6, colback=white, bottomrule=.15mm, leftrule=.75mm, titlerule=0mm, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomtitle=1mm, rightrule=.15mm, breakable, colbacktitle=quarto-callout-note-color!10!white, left=2mm]

@@

\subsection{Connecting optimisation success with its properties and
jellyfish
parameters}\label{connecting-optimisation-success-with-its-properties-and-jellyfish-parameters}

The optimisation properties (smoothness, squintability, and speed)
proposed in the Section~\ref{sec-theory} provide numerical measures to
characterise the difficulty of the optimisation for a given projection
pursuit problem. This example investigates how these measures, combined
with the jellyfish parameters, affect the success of the jellyfish
optimisers and how they can guide the decisions made in hyperparameter
tuning with the jellyfish optimiser. Simulations are conducted with
different jellyfish optimisers to obtain the proportion of successful
optimisation across a collection of scenarios, for which smoothness,
squintability, and speed are calculated. A generalised linear model is
used to link the proportion of success rate with the jellyfish
parameters and the optimisation properties.

In addition to the pipe-finding problem, also investigated is the
sine-wave finding problem in 6D and 8D spaces with six indices
considered: \texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC},
\texttt{TIC}, \texttt{spline}, and \texttt{stringy}. Combining with
different jellyfish optimiser parameters (\texttt{n\_jellies} and
\texttt{max\_tries}), a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, the
jellyfish optimiser is run fifty (50) times and summary statistics are
calculated on the best index value found across all 50 runs and the
proportion of runs that find a close (with a difference less than 0.05)
best index value (\texttt{P\_J\_hat}).

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance to the
optimal basis are calculated for each basis to fit a Gaussian process
model. For squintability, fifty random bases are sampled and
interpolated to the optimal basis with a step size of 0.005. The
interpolated bases are first binned with a bin width of 0.005 to average
the index values before feeding into a 4-parameter scaled logistic
function, estimated by non-linear least square, to obtain the
squintability measure.

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated for each case from the Gaussian process (variance, range,
smooth, and nugget) for smoothness and the scaled logistic function
(theta1 to theta4, and squint) for squintability. The column ``smooth''
and ``squint'' are used as the smoothness and squintability measure. The
low squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure. {[}TODO: the stringy squintability doens't make sense{]}

Table~\ref{tbl-mod-data} presents the processed data for modelling after
augmenting the jellyfish parameters (\texttt{n\_jellies} and
\texttt{max\_tries}) and the average time spent on each run
(\texttt{time}) for each case. A generalised linear model is fitted with
a binomial family and a logit link function to the data to investigate
the relationship between the proportion of jellyfish success and
jellyfish parameters and the optimisation properties. Pre-processing of
the data include 1) scale the jellyfish parameters (\texttt{n\_jellies}
and \texttt{max\_tries}) by 10 for interpretation, and 2) create a
binary variable \texttt{long\_time} to flag the cases with average time
spent on each run greater than 30 seconds. Table~\ref{tbl-mod-output}
presents the model output with the estimated coefficients. The model
suggests a positive relationship between the proportion of jellyfish
success and the jellyfish parameters, \texttt{n\_jellies} and
\texttt{max\_tries}, optimisation property parameter,
\texttt{smoothness} and \texttt{squintability}, and a negative
relationship with the average time spent on each run
(\texttt{long\_time}) and the data dimension (\texttt{d}). The variable
\texttt{squintability} and \texttt{d} are significant, suggesting their
importance over jellyfish parameters to the success of the optimisation.

\begingroup\fontsize{7}{9}\selectfont

@@

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (theta1 to theta4) for the pipe-finding and
sine-wave finding problems. The squint column is calculated as
\(\frac{\theta1 \theta2 \theta3}{4}\), as described in
Section~\ref{sec-squintability}. The ``smooth'' and ``squint'' column
represent the smoothness and squintability measures.}

@@

\endgroup{}

@@

\caption{\label{fig-idx-proj-dist}Index values against projection
distance for the 12 pipe/sine-wave finding problem after the binning
procedure during the estimation of the squintability measure. The index
value is averaged at each bin width of 0.005 and the \texttt{TIC} index
is scaled to 0-1 for comparison. When finding the sine-wave structure
using the \texttt{MIC} and \texttt{TIC} index, a convex curved is
observed, as opposed to the pipe-finding problem or the sine-wave
finding problem with the \texttt{dcor2d\_2}, \texttt{loess2d}, and
\texttt{splines2d} index. When finding the sine-wave with the stringy
index, the index shows an instantaneous jump to the optimum when close
to the best basis.}

@@

\caption{\label{tbl-mod-data}The first few rows of the datasets
processed for modelling. The smoothness and squintability variable are
uniquely characterised by the index function used and the data
dimension, and thus do not vary across \texttt{n\_jellies} and
\texttt{max\_tries}. The variable \texttt{P\_J\_hat}, and time are
calculated at each observation.}

@@

\begin{longtable}[t]{|>{}lrrr>{}r|}

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success (\texttt{P\_J\_hat}) on optimisation properties
(\texttt{smoothness}, \texttt{squintability}, and \texttt{d}) and
jellyfish parameters (\texttt{long\_time}, \texttt{n\_jellies}, and
\texttt{max\_tries}) from the generalised linear model with a binomial
family and a logit link function. The variable \texttt{long\_time} is
derived to flag the cases with average time spent on each run greater
than 30 seconds and the variable \texttt{n\_jellies} and
\texttt{max\_tries} are scaled by 10 for interpretation.}

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
intercept & -4.672 & 4.512 & -1.036 & 0.300\\
smoothness & 1.604 & 1.530 & 1.048 & 0.294\\
squintability & 7.062 & 2.210 & 3.195 & 0.001\\
d & -0.595 & 0.260 & -2.289 & 0.022\\
long\_time & -0.851 & 1.348 & -0.632 & 0.528\\
\addlinespace
n\_jellies & 0.230 & 0.131 & 1.754 & 0.080\\
max\_tries & 0.107 & 0.153 & 0.700 & 0.484\\
\bottomrule
